Reducing finitary inductive-inductive types to inductive types

### 2019-08-30 paper/universaliit.tex
some typos in the universal IIT part of the paper

\subsubsection{Untyped weakening}

@@

  \subsubsection{Untyped substitution}
  We define by induction the specific substitution of one variable of the context:
  % & \rlap{(1) Substitution calculus} \\[0.5em]
\begin{alignat*}{5}
  & \blank[\blank := \blank] && : \Tyu\ra\N\ra\Tmu\ra\Tyu \\
  & \blank[\blank := \blank] && : \Tmu\ra\N\ra\Tmu\ra\Tmu
  % \\
  % & \blank[\blank := \blank] && : \N\ra\N\ra\Tmu\ra\Tmu
  \end{alignat*}
  This is enough to define the well-typed judgments: indeed, the typing rule of application
  involves only a unary substitution.

@@

  \subsubsection{Typing judgments}
  The typing judgments are defined as the following inductive datatype indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \Gamma\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \Gamma\vdash )  \ra  (\Gamma\vdash A)  \ra
  \Gamma\extu A \vdash \\
  & \consw && :
    (\Delta \vdash) \ra
    (\Gamma\vdash \sigma \Rightarrow\Delta)\ra
    ( \Delta \vdash A) \ra
    ( \Gamma \vdash t \in A[\sigma]) \ra
    \Gamma \vdash t \consu \sigma \Rightarrow \Delta\extu A
   \\
  & \varw  && : (\Gamma \vdash n \invar A) \ra \Gamma \vdash \varu n \in A \\
  & \vzw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra\Gamma \extu  A \vdash 0 \invar \wku \, A \\
  & \vsw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra(\Gamma \vdash n \invar A) \ra (\Gamma \vdash B) \ra \Gamma \extu  B
  \vdash S\,n \invar \wku \, A \\
  & \rlap{(2) Sorts and operators} \\[0.5em]
  & \Uw && : (\Gamma \vdash)\ra \Gamma \vdash \Uu \\
  & \Elw && : (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra\Gamma \vdash \Elu\,a \\
  & \rlap{(3) Parameters} \\[0.5em]
  & \Piw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
  \Elu\,a \vdash B)\ra\Gamma \vdash \Piu \, a \, B \\
  & \appw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
    \Elu\,a \vdash B)\ra
    \\
    & &&
    (\Gamma \vdash t \in \Piu \, a \, B )
    \ra
    (\Gamma \vdash u \in \Elu\,a)
    \ra
    \Gamma \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t) \ra
    \Gamma \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t) \ra
    \\
    & &&
    (\Gamma \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
    \\
  & \rlap{(5) Infinitary parameters} \\[0.5em]
  & \Piiw && :
    (T:\Set)\ra(A : T\ra \Tmu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t\in \Uu)
    \\ &&& \ra
    \Gamma \vdash \Pimu \,T\,A \in \Uu
  \\
  & \appiw && :
    (T:\Set)\ra(A : T\ra \Tmu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t\in\Uu)
    \\
    & &&
    \ra
    (\Gamma \vdash t \in \Piiu \,T\, A)
    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
\end{alignat*}
Note that there is possibility of redundancy in the arguments of
the constructors. Here, we are rather ``{paranoid}'', so that we get more inductive
hypotheses when performing recursion.

@@

Then, we show that identity substitutions is neutral, for well-typed types,

@@

\subsubsection{Uniqueness of typing and uniqueness of types}
  We define what it means for a type $A$ to be a proposition as follows:

  \[
    \isaprop\,A := (a : A) \ra (a' : A) \ra a = a'
  \]
  It states that there is at most one inhabitant.
We show by recursion that each of these typing judgments is unique in the
following sense:
\begin{align*}
    \isp{\Conw}
  : \isaprop\, (\Gamma\vdash)
 & &
    \isp{\Tmw}
  : \isaprop\, (\Gamma\vdash t \in A)
    \\
    \isp{\Tyw}
  : \isaprop\, (\Gamma\vdash A)
    &&
    \isp{\wf\Var}
  : \isaprop\, (\Gamma\vdash x \invar A)
  \\
  &
        \isp{\wf\Sub}
  : \isaprop\, (\Gamma\vdash \sigma \in \Delta)
  \end{align*}
Actually, the application and successor cases require to show uniqueness of
types in a given context, before.
Let us consider for instance the application constructor $\appw$: it provides a
type $B$ such that the final type is $C=B[0 := u]$. Even if $C$ is known a
priori, there may be another $B$ for which $B[0 := u] = C$, possibly leading to
many proofs that $t\appu u$ has type $C$. Uniqueness of types
solves this issue, as $B$ is then uniquely determined by the type $\Piu\,A\,B$ of $t$.
More formally, we show by recursion:
\begin{alignat*}{5}
  &
  \Tmw{=}\Ty & &:
  (\Gamma \vdash t \in A) \ra
  (\Gamma \vdash t \in B) \ra A = B
  \\
  &
  \wf\Var{=}\Ty & &:
  (\Gamma \vdash x \invar A) \ra
  (\Gamma \vdash x \invar B) \ra A = B
  \end{alignat*}

@@

Given a model $M$ of the QIIT, we define the functional relation enjoyed by the initial
morphism $\rec : S\to M$ by recursion on the typing judgments.
% As these judgments are propositions, we will sometimes
If $\Gamma$ is a context in $S$ and $\model{\Gamma}$ is a semantic
context (i.e. a context of the model $M$), we want to define a type $\Gamma\sim \model{\Gamma}$
equivalent to $\rec(\Gamma)=\model{\Gamma}$. Of course, at this
stage, $\rec$ is not available yet since the whole point of defining this
relation is to construct $\rec$ in the end. For a type $A$ in a context $\Gamma$,
it is natural to define the relation with a semantic type $\model{A}$ in a
semantic context $\model{\Gamma}$ such that it is equivalent to  $(\rec\,\Gamma =
\model{\Gamma})\times (\rec\,A = \model{A})$ (note that the first equality should
be used to transport the right hand side of the second one).
The discussion is similar for terms.

The ultimate goal is to prove that
$\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
$\sum_{\model{\Gamma}}({\model{A}:\model{\Ty}\,\model{\Gamma}})\times(A\sim\model{A})$
are contractible. The first step consists in showing that these types are propositions.
% We explain below that this requires UIP.
Actually, we take a different approach for the relation on types
(and terms) which yields a more concise definition of the relation.
The adapted goal, for types, will be to prove that
$\sum_{\model{A}}A\sim\model{A}$ is a proposition, and
is inhabited if if $\Gamma \sim \model{\Gamma}$ is.
 This means that in the definition of the relation, we assume
that contexts are already related, so that we don't need to enforce them to be,
as it is the case in the original approach. We give the two versions of the universe
case below, to explain the differences.

>>>>

\subsubsection{Weakening preserves typing}

@@

  \subsubsection{Substitution preserves typing}
  We define by induction the specific substitution of one variable of the context:
  % & \rlap{(1) Substitution calculus} \\[0.5em]
\begin{alignat*}{5}
  & \blank[\blank := \blank] && : \Tyu\ra\N\ra\Tmu\ra\Tyu \\
  & \blank[\blank := \blank] && : \Tmu\ra\N\ra\Tmu\ra\Tmu
  % \\
  % & \blank[\blank := \blank] && : \N\ra\N\ra\Tmu\ra\Tmu
  \end{alignat*}
  This is enough to define the well-typed judgments: indeed, the typing rule of application
  involves only a unary substitution.

@@

  \subsubsection{Typing judgments}
  The typing judgments are defined as the following inductive datatype indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \Gamma\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \Gamma\vdash )  \ra  (\Gamma\vdash A)  \ra
  \Gamma\extu A \vdash \\
  & \consw && :
    (\Delta \vdash) \ra
    (\Gamma\vdash \sigma \Rightarrow\Delta)\ra
    ( \Delta \vdash A) \ra
    ( \Gamma \vdash t \in A[\sigma]) \ra
    \Gamma \vdash t \consu \sigma \Rightarrow \Delta\extu A
   \\
  & \varw  && : (\Gamma \vdash n \invar A) \ra \Gamma \vdash \varu n \in A \\
  & \vzw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra\Gamma \extu  A \vdash 0 \invar \wku \, A \\
  & \vsw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra(\Gamma \vdash n \invar A) \ra (\Gamma \vdash B) \ra \Gamma \extu  B
  \vdash S\,n \invar \wku \, A \\
  & \rlap{(2) Sorts and operators} \\[0.5em]
  & \Uw && : (\Gamma \vdash)\ra \Gamma \vdash \Uu \\
  & \Elw && : (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra\Gamma \vdash \Elu\,a \\
  & \rlap{(3) Parameters} \\[0.5em]
  & \Piw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
  \Elu\,a \vdash B)\ra\Gamma \vdash \Piu \, a \, B \\
  & \appw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
    \Elu\,a \vdash B)\ra
    \\
    & &&
    (\Gamma \vdash t \in \Piu \, a \, B )
    \ra
    (\Gamma \vdash u \in \Elu\,a)
    \ra
    \Gamma \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Metatheoretic parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t) \ra
    \Gamma \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t) \ra
    \\
    & &&
    (\Gamma \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
    \\
  & \rlap{(5) Infinitary parameters} \\[0.5em]
  & \Piiw && :
    (T:\Set)\ra(A : T\ra \Tmu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t\in \Uu)
    \\ &&& \ra
    \Gamma \vdash \Pimu \,T\,A \in \Uu
  \\
  & \appiw && :
    (T:\Set)\ra(A : T\ra \Tmu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t\in\Uu)
    \\
    & &&
    \ra
    (\Gamma \vdash t \in \Piiu \,T\, A)
    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
\end{alignat*}
There is possibility of redundancy in the arguments of
the constructors. Here, we are rather ``{paranoid}'', so that we get more inductive
hypotheses when performing recursion.

@@

Then, we show that identity substitutions are neutral, for well-formed types,

@@

\subsubsection{Uniqueness of typing and uniqueness of types}
  We define what it means for a type $T$ (of the meta-theory) to be a proposition as follows:

  \[
    \isaprop\,T := (a : T) \ra (a' : T) \ra a = a'
  \]
  Thus, a type is a proposition if there is at most
  one inhabitant.
We show by recursion that each of these typing judgments is unique in the
following sense:
\begin{align*}
    \isp{\Conw}
  : \isaprop\, (\Gamma\vdash)
 & &
    \isp{\Tmw}
  : \isaprop\, (\Gamma\vdash t \in A)
    \\
    \isp{\Tyw}
  : \isaprop\, (\Gamma\vdash A)
    &&
    \isp{\wf\Var}
  : \isaprop\, (\Gamma\vdash x \invar A)
  \\
  &
        \isp{\wf\Sub}
  : \isaprop\, (\Gamma\vdash \sigma \Rightarrow \Delta)
  \end{align*}
Actually, the application and successor cases require to show uniqueness of
types in a given context, before.
Let us consider for instance the application constructor $\appw$: it provides a
type $B$ such that the final type is $C=B[0 := u]$. Even if $C$ is known a
priori, there may be another $B$ for which $B[0 := u] = C$, possibly leading to
many proofs that $t\appu u$ has type $C$. Uniqueness of types
solves this issue, as $B$ is then uniquely determined by the type $\Piu\,A\,B$ of $t$.
More formally, we show by recursion:
\begin{alignat*}{5}
  &
  \Tmw{=}\Ty & &:
  (\Gamma \vdash t \in A) \ra
  (\Gamma \vdash t \in B) \ra A = B
  \\
  &
  \wf\Var{=}\Ty & &:
  (\Gamma \vdash x \invar A) \ra
  (\Gamma \vdash x \invar B) \ra A = B
  \end{alignat*}

@@

Given a model $M$ of the QIIT, we define the functional relation enjoyed by the initial
morphism $\rec : S\to M$ by recursion on the typing judgments.
% As these judgments are propositions, we will sometimes
If $\Gamma$ is a context in $S$ and $\model{\Gamma}$ is a semantic
context (i.e. a context of the model $M$), we want to define a type $\Gamma\sim \model{\Gamma}$
equivalent to $\rec(\Gamma)=\model{\Gamma}$. Of course, at this
stage, $\rec$ is not available yet since the whole point of defining this
relation is to construct $\rec$ in the end. For a type $A$ in a context $\Gamma$,
it is natural to define the relation with a semantic type $\model{A}$ in a
semantic context $\model{\Gamma}$ such that it is equivalent to  $(\rec\,\Gamma =
\model{\Gamma})\times (\rec\,A = \model{A})$ (note that the first equality is
required for the second one to be well-typed).
The discussion is similar for terms.

The ultimate goal is to prove that
$\sum_{\model{\Gamma}}\Gamma\sim\model{\Gamma}$ and
$\sum_{\model{\Gamma}}({\model{A}:\model{\Ty}\,\model{\Gamma}})\times(A\sim\model{A})$
are contractible, i.e., have a unique inhabitant. The first step consists in showing that these types are propositions.
% We explain below that this requires UIP.
Actually, we take a different approach for the relation on types
(and terms) which yields a more concise definition of the relation.
The adapted goal, for types, will be to prove that
$\sum_{\model{A}}A\sim\model{A}$ is a proposition, and
is inhabited if if $\Gamma \sim \model{\Gamma}$ is.
 This means that in the definition of the relation, we assume
that contexts are already related, so that we don't need to enforce them to be,
as it is the case in the original approach. We give the two versions of the universe
case below, to explain the differences.
